---
title: "Identifying likely dephosphorylated residues and specific phosphatase activity"
author: "Nataliya Trushina"
date: "`r Sys.Date()`"
output: 
  word_document:
    toc: true
    reference_docx: "../Rmarkdown_template.docx"
---

# Introduction and model choice

As phosphatase activity prediction proxy, we used the DTL-DephosSite model[1], a deep learning tool for predicting protein dephosphorylation sites. It uses transfer learning based on kinase-specific phosphorylation site prediction, capturing shared sequence patterns and biochemical features between phosphorylation and dephosphorylation. The lower network layers are pre-trained on phosphorylation data, while the upper layers are trained specifically on dephosphorylation to adapt for phosphatase predictions.

Earlier models, such as those in [2], [3], and [4], introduced computational approaches for dephosphorylation site prediction, but their servers are no longer available.


1. Chaudhari, M., Thapa, N., Ismail, H., Chopade, S., Caragea, D., Köhn, M., ... & Kc, D. B. (2021). DTL-DephosSite: Deep transfer learning based approach to predict dephosphorylation sites. Frontiers in cell and developmental biology, 9, 662983. https://www.frontiersin.org/journals/cell-and-developmental-biology/articles/10.3389/fcell.2021.662983/full

2. Wang, X., Yan, R., & Song, J. (2016). DephosSite: a machine learning approach for discovering phosphotase-specific dephosphorylation sites. Scientific Reports, 6(1), 23510. https://pmc.ncbi.nlm.nih.gov/articles/PMC4802303/

3. Jia, C., He, W., & Zou, Q. (2017). DephosSitePred: a high accuracy predictor for protein dephosphorylation sites. Combinatorial Chemistry & High Throughput Screening, 20(2), 153-157. https://pubmed.ncbi.nlm.nih.gov/28031011/

4.Wu, Z., Lu, M., & Li, T. (2014). Prediction of substrate sites for protein phosphatases 1B, SHP-1, and SHP-2 based on sequence features. Amino Acids, 46, 1919-1928.
https://link.springer.com/article/10.1007/s00726-014-1739-6

---

# Step 1 - Preprocessing: Prepare dataset in the required format

Use the output from `phosphoproteomics_PEAKSonline.R` (https://github.com/Department-of-Neurobiology/Meta-Proteomics). Create a `config.R` file to save the configureations for all the pipeline scripts.

```{r}
path_to_file <- "Dataset_processesed_annotated_for_iGPS_and_KEA2_full.xlsx"
condition <- "arsenite_decreased"
col_control <- "#4f4f4f"
col_exp <- "#ED6B4D"
prob_cut_off <- 0.5
writeLines(condition, "config.txt")  # For running Python script with the same condition

```

Run `00_Phosphatase_analysis_preprocessing.R`:

1. Start with a list of peptides showing significantly **lower** phosphorylation in the experimental condition (H2O2 or arsenite) versus control.

2. Extract 31-residue windows around each residue of interest:
     - **15 residues upstream** and **15 residues downstream**.
     
3. Save the sequences in FASTA format:
     - Example:
       ```
       >sp|Q63788|P85B_RAT%583%615
       RKINEWLGIKNETEDQYSLMEDEDALPHHEERT"
       ```
4. Split by residue type:
   - Create two separate FASTA files:
     - One for **S/T residues** (serine/threonine).
     - One for **Y residues** (tyrosine).


```{r}
source("00_Phosphatase_analysis_preprocessing.R")

```

---

# Step 2 - Modelling: Run the DTL-DephosphoSite models

Run `01_model_apply.py`:

1. Both models are run directly. 

2. Y-model was trained on shorter sequences so it requires a different input for the window size.

3. The prediction score is given as a probability. The sensitivity and specificity of the models was calculated based on the cut-off of 0.5. 

---

# Step 3 - Postprocessing: Extend the analysis

1. Count the fractions of predicted sites to negative.

2. Predict specific phosphatases:
   - **DEPOD** ([http://depod.bioss.uni-freiburg.de/](http://depod.bioss.uni-freiburg.de/)): Provides known phosphatases for specific proteins and residues. Download "Inter-relations between human phosphatases, kinases and their common substrates" dataset.

3. Translate human to rat ortologs and check the possible phosphatases and substrates in the proteins with decrease phosphorylation upon treatment.

*Note: DEPOD database is relatively small, ortolog translation does not achieve better matching*

4.The proteins of interest have to be checked to see if the decreased phosphosites were possibly the ortologous site in the DEPOD database.

---

# Conclusion

Residues predicted as "dephosphorylated" suggest active phosphatase activity.


